---
title: "Appendix 1"
author: "Eslava, Ada"
date: "March 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<style>
body {
text-align: justify}
</style>

###**Description:**

This Appendix contains the code to:
1) Generate maps of the surveyed sites 
2) Generate extirpations vs introductions spatial map
3) Explore features of the data
4) Compute the species' accumulation curves & estimated gamma diversity

<p>&nbsp;</p>


#### **Libraries:**
```{r, echo = TRUE, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)  
library(rworldmap) 
library(maps)
library(maptools)
library(grid)
library(vegan)
library(kableExtra)
library(gridExtra)
library(scales)
library(GeoRange)
library(sp)
library(sf)
library(hablar)
library(magrittr)
library(legendMap)
library(iNEXT.3D)
```
<p>&nbsp;</p>
#### **Read data: **
```{r, echo=TRUE, warning=FALSE, message=FALSE}
myd <- getwd()
rm(list=ls())
path_lists <- "C:/Users/afe1/OneDrive - University of St Andrews/PHD/ThesisChapterMexico_I/TemporalChange_MexicanFish_C2/Lists"
load(paste0(path_lists, "/l67.RData"))   # data.frame of formatted metacom data matrices 67
load(paste0(path_lists, "/l53.RData"))   # data.frame of formatted metacom data matrices 53
load(paste0(path_lists, "/NDT67.RData")) # rawdata sheet 67
```
<p>&nbsp;</p>

#### **Site Types: **
```{r, echo=TRUE, warning=FALSE, message=FALSE}
length(unique(NDT67$SiteType)) # 13
NDT67 %>% group_by(SiteType) %>% summarise(n())
```
<p>&nbsp;</p>
#### **Surveyed area: **
Retrieve area covered in KM2 & plot:
```{r, echo=TRUE}
# Convex hulls =================================================================
# This chunck uses the 67 site subset
range(NDT67$Latitude)
range(NDT67$Longitude)
dt_coord <- NDT67 %>% select(Longitude, Latitude) %>% distinct() %>%
  st_as_sf(., coords = c('Longitude', 'Latitude'), crs = 4326) %>% st_union()

centroid <- dt_coord %>% st_convex_hull() %>% st_centroid() %>% unlist # get centroid
centroid

area <- st_transform(dt_coord, st_crs("+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=km +no_defs")) %>%
  st_convex_hull() %>% st_area()  ##get area in sq km
area

angle2dec <- function(angle) {
  angle <- as.character(angle)
  x <- do.call(rbind, strsplit(angle, split=' '))
  x <- apply(x, 1L, function(y) {
    y <- as.numeric(y)
    y[1] + y[2]/60 + y[3]/3600
  })
  return(x)
}

world_map <- map_data('world', c('Mexico'), xlim=c(-115, -90), ylim=c(14,28))
ggplot(world_map) +
  geom_polygon(aes(x=long, y=lat, group=group), color='grey', fill='transparent') +
  geom_point(data = NDT67, aes(x = Longitude, y = Latitude), size = 2, shape = 21, colour = 'blue') +
  geom_point(aes(x = centroid[1], y = centroid[2]), size = 3, colour = 'green') +
  geom_sf(data = dt_coord %>% st_convex_hull(), colour = 'blue', fill = 'transparent') +
  labs(x = NULL, y = NULL) +
  theme_minimal() # basic map

```

```{r, echo=F, warning=FALSE, message=FALSE}

#PREV TITLE: #### **Area of basins: **

#basins <- split(NDT67, f=NDT67$DrainageBasinE)
#basins_pointsKm2 <- lapply(basins, function(x) {SpatialPoints(cbind(x$Longitude,x$Latitude), proj4string=CRS('+proj=longlat +ellps=WGS84 +datum=WGS84 #+no_defs')) %>% sp::spTransform(., CRS("+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=km +no_defs"))})
#basins_areaKm2 <- as.data.frame(do.call(rbind, lapply(basins_pointsKm2, function(x) {gArea(gConvexHull(x))})))
#names(basins_areaKm2) <- "areakm2"

# Remove those basins with 1 or 2 sites only.
#round(basins_areaKm2, 0) %>%
#  filter(areakm2>0) %>% # Basins with one or two sites only
#  kbl() %>%
#  kable_styling() # 9 basins

# NOTE: rm, dependency of RGEOS, which is DEPRECATED
```

```{r, echo=F, results=FALSE, message=FALSE, warning=FALSE}
#PREV TITLE:#### **Retrieve other information: **
# These data contains all sites. Sites that could not be included in analyses are kept in this subset.


#length(unique(NDT67$DrainageBasinE))                             # 19 basins
#n <- NDT67 %>% group_by(DrainageBasinE) %>% summarise(count=n()) # 1 to 10 sites/basin
#nn <- n %>% group_by(count) %>% summarise(countII=n())           # 9 basins with more than 3 sites.

NDT67$CountNat2005 <- gsub("-",0, NDT67$CountNat2005)
NDT67$CountNonNat2005 <- gsub("-",0, NDT67$CountNonNat2005)
NDT67 <- NDT67 %>% retype()

range(NDT67$CountHistNatCatalog, na.rm=TRUE)                     # 1 to 17
range(NDT67$CountHistNatBroad, na.rm=TRUE)                       # 1 to 17

sort(unique(NDT67$HistoricalExotics, na.rm=TRUE))                # 4 species reported as exotics in the historical period
range(NDT67$CountNat2005, na.rm=TRUE)                            # 0 to 9
range(NDT67$CountNonNat2005)                                     # 0 to 6

sum(NDT67$CountNat2005==0)                                       # 21
sum(NDT67$CountNonNat2005==0)                                    # 30
sum(NDT67$`RatioFound/Expected`==1)                              # 13 (preserved native assemblage)
sum(NDT67$`RatioFound/Expected`[NDT67$CountNonNat2005==0]==1)    # 7 (preserved native + no nonative)


sort(unique(NDT67$CatalogNumbers))
sort(unique(NDT67$Collectors))
sort(unique(NDT67$DateCollected))                                # 1891 to 1986

# Notes & Comments
sort(unique(NDT67$Anectodic))
sort(unique(NDT67$Comments))

#NOTES:
#The dataset contains historical and contemporary data from 67 sites in 19 drainage basins. These sites represent different bodies of water, both lentic & #lotic. The number of sites x basin ranges from 1 (lakes) to 10. Ten of the basins are represented by only 1 or 2 sites, on the other nine by more than 3.
#In the historical period, the richness of native species ranges from 1 to 17 species. The richness of non-native fish is mostly 0 but there are 4 #non-native fish that appear in a few sites.
#In the contemporary period, the richness of native fish species ranges from 0 to 9 and the richness of non-natives ranges from 0 to 6. There are no native #fish in 7 sites, and no non-native fish in 30 sites including those where no fish life was found. The native assemblage is preserved in 13 sites, 7 of #which contain no introduced fishes.
#The historical collections are reconstructed with catalog records from the University of Michigan Museum of Zoology and the United States National Museum #(Smithsonian). These records are mostly reported by Miller, but also other authors. The date of these collections ranges from 1891 to 1986 and therefore #the historical period includes data that was collected in different time windows over a period of ~ 100 years.

```
<p>&nbsp;</p>
#### **Mapping local extirpations and introductions:**

```{r}
plot_dir <- "C:/Users/afe1/OneDrive - University of St Andrews/PHD/ThesisChapterMexico_I/TemporalChange_MexicanFish_C2/Plots" # Dir to save main plots

# Map Status ===================================================================
# This chunck uses the 67 site subset

# COLOR KEY:
# Black == no fish
# Dark orange == no native fish (completely new assemblage)
# Dark green == completely intact assemblage (all natives found and no non-natives)

# Yellow orange == extirpations occuring
# Reddish purple == extirpations and introductions occurring
# Blue == introductions occuring


no_fish <- na.omit(NDT67$SiteNameE[(NDT67$CountNat2005+NDT67$CountNonNat2005)==0])   # 14
new_assemblage <- NDT67$SiteNameE[NDT67$CountNat2005==0 & NDT67$CountNonNat2005>0]   # 7
intact <- NDT67$SiteNameE[NDT67$CountNonNat2005==0 & NDT67$`RatioFound/Expected`==1] # 7
extirpationsonly <- NDT67$SiteNameE[NDT67$CountHistNatBroad>NDT67$CountNat2005 & NDT67$CountNat2005>0 & NDT67$CountNonNat2005==0] # 9
introductionsonly <- NDT67$SiteNameE[NDT67$CountHistNatBroad==NDT67$CountNat2005 & NDT67$CountNonNat2005>0] # 6

67 - (14+7+7+9+6) # 24 (both)

ext_intro <- NDT67$SiteNameE[NDT67$CountHistNatBroad>NDT67$CountNat2005 & NDT67$CountNat2005>0 & NDT67$CountNonNat2005>0] # 24, OK

# Create plotting dataframe: -
df <- data.frame("Site"=NDT67$SiteNameE, "Basin"=NDT67$DrainageBasinE, "Latitude"=NDT67$Latitude,"Longitude"=NDT67$Longitude, "Status"=rep(NA, nrow(NDT67))) 

df$Status <- ifelse(df$Site %in% no_fish, "No fish found (14)", df$Status) # Black
df$Status <- ifelse(df$Site %in% new_assemblage, "New assemblage (100% non-natives) (7)", df$Status) # Dark blue
df$Status <- ifelse(df$Site %in% intact, "Intact assemblage (7)", df$Status) # Green
df$Status <- ifelse(df$Site %in% extirpationsonly, "Extirpations but no introductions (9)", df$Status) # Dark blue
df$Status <- ifelse(df$Site %in% introductionsonly, "Introductions but no extirpations (6)", df$Status) # Green
df$Status <- ifelse(is.na(df$Status), "Introductions & extirpations (24)", df$Status) # Both introductions and extirpations
sum(is.na(df$Status)) # 0, OK

df$StatusII <- ifelse(df$Status=="Intact assemblage (7)", "Intact assemblage (7)", "Modified assemblage (60)") 
sum(is.na(df$StatusII)) # 0, OK

# Mexico Map:
world_map <- map_data('world', c('Mexico'), xlim=c(-115, -90), ylim=c(14,28))
world_mapII <- world_map[world_map$long > -107 & world_map$long < -97,]
world_mapII <- world_mapII[world_mapII$lat < 28 & world_mapII$lat > 16.5,]

g0<-ggplot()+coord_fixed()+xlab("")+ylab("")
g0<-g0+geom_polygon(data=world_map, aes(x=long, y=lat, group=group), colour="black", fill="gray66")
g0<-g0+theme_bw()
g0<-g0 + legendMap::scale_bar(lon = -115, lat = 15, 
              distance_lon = 250, distance_lat = 50, distance_legend = 100, 
              dist_unit = "km", orientation = TRUE,legend_size = 6,
              arrow_length = 200, arrow_distance = 200, arrow_north_size = 6)
g0<-g0 + geom_rect(aes(xmin = -107, xmax = -97, ymin = 16.5, ymax = 28), color = "black", fill = NA)
g0 <- g0 + theme(legend.text = element_text(size=18),
                  #legend.title = element_text(size=14),
                  legend.position = "left",
                  #plot.title = element_text(size=16, face="bold", hjust=0.5, vjust=1),
                  axis.text.x = element_text(size=18),
                  axis.text.y = element_text(size=18))

g1<-ggplot()+coord_fixed()+xlab("")+ylab("")
g1<-g1+geom_polygon(data=world_mapII, aes(x=long, y=lat, group=group), colour="black", fill="gray66")
g1<-g1+theme_bw()
g1<-g1 + legendMap::scale_bar(lon = -107, lat = 17, 
              distance_lon = 100, distance_lat = 30, distance_legend = 100, legend_size = 4,
              dist_unit = "km", orientation = FALSE)
g1<-g1+labs(title="", fill="")
g1 # basic canvas


##Panel A: ---------------------------------------------------------------------
g1A <- g1+geom_point(data = df,  # Add and plot site data
                     aes(x = Longitude, y = Latitude, 
                     fill = Status), shape=21, size=5, alpha=0.9)
(g1A <- g1A + theme(legend.text = element_text(size=16),
                  #legend.title = element_text(size=14),
                  legend.position = "right",
                  #plot.title = element_text(size=16, face="bold", hjust=0.5, vjust=1),
                  axis.text.x = element_text(size=16),
                  axis.text.y = element_text(size=16)))
g1A <- g1A+scale_fill_manual(values = c("No fish found (14)" = "grey18",
                                     "New assemblage (100% non-natives) (7)"= "#D55E00",
                                     "Intact assemblage (7)"="#009E73",
                                     "Extirpations but no introductions (9)"="#56B4E9",
                                     "Introductions but no extirpations (6)"="#E69F00",
                                     "Introductions & extirpations (24)"="#CC79A7"))


##Panel B: ---------------------------------------------------------------------
g1B<-g1+geom_point(data = df,  # Add and plot site data
                   aes(x = Longitude, y = Latitude, 
                       fill = StatusII), shape=21, size=5, alpha=0.9)
(g1B <- g1B + theme(legend.text = element_text(size=16),
                  #legend.title = element_text(size=14),
                  legend.position = "right",
                  #plot.title = element_text(size=16, face="bold", hjust=0.5, vjust=1),
                  axis.text.x = element_text(size=16),
                  axis.text.y = element_text(size=16)))
g1B<-g1B+scale_fill_manual(values = c("Intact assemblage (7)"="#009E73",
                                     "Modified assemblage (60)"="grey18"))

MexicoMap <- g0
MexicoMapZoomA <- g1A
MexicoMapZoomB <- g1B

ggsave(MexicoMap, file= paste0(plot_dir, "/Chapter5/MexicoMap.jpg"), width = 12, height = 8)
ggsave(MexicoMapZoomA, file= paste0(plot_dir, "/Chapter5/MexicoMapZoomA.jpg"), width = 12, height = 8)
ggsave(MexicoMapZoomB, file= paste0(plot_dir, "/Chapter5/MexicoMapZoomB.jpg"), width = 12, height = 8)

# Proportions of the total:
14/67*100
7/67*100
9/67*100
24/67*100
6/67*100
```
<p>&nbsp;</p>
#### **Richness: **

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Singletons, doubletons, etc: =================================================
# This chunck uses the 67 site subset 

HNC <- subset(l67, l67$Period=="HNC")
HNB <- subset(l67, l67$Period=="HNB")
ContAll <- subset(l67, l67$Period=="ContAll")
#HNC <- HNC[!rowSums(HNC[,c(4:ncol(HNC))])==0,]
#HNB <- HNB[!rowSums(HNB[,c(4:ncol(HNB))])==0,]
#ContAllII <- ContAll[!rowSums(ContAll[,c(4:ncol(ContAll))])==0,]
listComms <- list(HNC, HNB, ContAll)
listCommsL <- lapply(listComms, function(x){gather(x, key="Species", value="S", -c(1:3))})
S <- lapply(listCommsL, function(x) {x <- x %>% group_by(SiteNameE) %>% summarise(S=sum(S))})

# Nº sites with S=1, S=2 or S>3:
S1 <- lapply(S, function(x) {x <- subset(x, x$S==1)}) # HC: 19, HB: 7 & Cont: 4
S2 <- lapply(S, function(x) {x <- subset(x, x$S==2)}) # HC: 2, HB: 8 & Cont: 14
S3 <- lapply(S, function(x) {x <- subset(x, x$S>3)})  # HC: 34, HB: 44, Cont: 27
```
<p>&nbsp;</p>
#### **Species accumulation curves:**

```{r, echo=TRUE, results=TRUE, fig.height=7, fig.width=6}
# Species accumulation curves: =================================================
# This chunck uses both the 53 and the 67 site subset 

list_mat67 <- list("HNC"=as.matrix(HNC[,c(4:ncol(HNC))]), "HNB"=as.matrix(HNB[,c(4:ncol(HNB))]),"ContAll"=as.matrix(ContAll[,c(4:ncol(ContAll))]))

HNC53 <- subset(l53, l53$Period=="HNC")
HNB53 <- subset(l53, l53$Period=="HNB")
ContAll53 <- subset(l53, l53$Period=="ContAll")
list_mat53 <- list("HNC"=as.matrix(HNC53[,c(4:ncol(HNC53))]), "HNB"=as.matrix(HNB53[,c(4:ncol(HNB53))]),"ContAll"=as.matrix(ContAll53[,c(4:ncol(ContAll53))]))


function_speaccum_graphs <- function(l){
  listSpecaccum <- lapply(l, function(x) {x <- specaccum(x, method="random", permutations=1000)})
  listSpecaccum <- lapply(listSpecaccum, function(x) {x <- data.frame(cbind(x$sites, x$richness, x$sd))
  names(x) <-c("Sites", "Richness", "Sd");x})
  names(listSpecaccum) <- c("a) Historical Conservative", "b) Historical Broad", "c) Contemporary Natives + Exotics")

  list_plots <- list()
  
  for (i in 1:length(listSpecaccum)){
  dt <- listSpecaccum[[i]]
  x <- names(listSpecaccum)[[i]]
  SpCurves <- ggplot(data=dt, aes(y=Richness, x=Sites)) +
  geom_line() +
  geom_ribbon(aes(ymin=(Richness-2*Sd), ymax=(Richness+2*Sd)),alpha=0.2)+ # Does a bit of a weird thing for low levels of richness
  ylim(c(0,100))+
  scale_x_continuous(breaks=c(0,20,40,60,80,100), 
                   labels=c("0","20","40","60","80","100"))+
  labs(title=x)+
  theme_classic()+
  theme(plot.title = element_text(size = 16, face = "bold"),
    legend.title=element_text(size=14), 
    legend.text=element_text(size=12))
  list_plots[[i]] <- SpCurves
}
  SpecaccumP <- grid.arrange(grobs=list_plots, nrow=2)
  return(SpecaccumP)
  
}


speaccum67 <- function_speaccum_graphs(list_mat67)
speaccum53 <- function_speaccum_graphs(list_mat53)


ggsave(speaccum67, file=paste0(plot_dir, "/Chapter5/speaccum67.jpg"), width=8, height=8)
ggsave(speaccum53, file=paste0(plot_dir, "/Chapter5/speaccum53.jpg"), width=8, height=8)

```
<p>&nbsp;</p>

#### **Sample coverage (iNEXT):**
```{r, echo=TRUE, results=TRUE, fig.height=7, fig.width=6}
# iNEXT : ======================================================================
# This chunck uses both the 53 and the 67 site subset
list_mat53t <- lapply(list_mat53, function(x) {t(x)})
list_mat67t <- lapply(list_mat67, function(x) {t(x)})

iNEXT53 <- lapply(list_mat53t, function(x) {iNEXT3D(data = x, diversity = "TD",
                                           q = 0, datatype = "incidence_raw", nboot = 100)})

iNEXT67 <- lapply(list_mat67t, function(x) {iNEXT3D(data = x, diversity = "TD",
                                            q = 0, datatype = "incidence_raw", nboot = 100)})


iNEXT53plot <- lapply(iNEXT53, function(x) {x[["TDiNextEst"]][["size_based"]]})
iNEXT53plot <-Map(cbind, iNEXT53plot, "Period"=names(iNEXT53plot))
iNEXT53plot <- as.data.frame(do.call(rbind, iNEXT53plot))
iNEXT53plot$Period <- factor(iNEXT53plot$Period, levels = c("HNC", "HNB" , "ContAll"),
                             labels = c("a) Historical Conservative", "b) Historical Broad", "c) Contemporary Natives + Exotics"))


iNEXT67plot <- lapply(iNEXT67, function(x) {x[["TDiNextEst"]][["size_based"]]})
iNEXT67plot <- Map(cbind, iNEXT67plot, "Period"=names(iNEXT67plot))
iNEXT67plot <- as.data.frame(do.call(rbind, iNEXT67plot))
iNEXT67plot$Period <- factor(iNEXT67plot$Period, levels = c("HNC", "HNB" , "ContAll"),
                             labels = c("a) Historical Conservative", "b) Historical Broad", "c) Contemporary Natives + Exotics"))


size_vs_coverage <- function(dt_plot){

  dt_plot_ref <- subset(dt_plot, dt_plot$Method=="Observed")
  dt_plot_r <- subset(dt_plot, dt_plot$Method=="Rarefaction")
  dt_plot_e <- subset(dt_plot, dt_plot$Method=="Extrapolation")
  
   p <- ggplot(data=dt_plot, aes(x=mT, y=SC, color=Period)) +
    labs(y="sample coverage", x="size")+
   geom_point(data=dt_plot_ref, aes(x=mT, y=SC))+
   geom_line(data=dt_plot_r, aes(x=mT, y=SC), linetype="solid", alpha=0.4)+
   geom_line(data=dt_plot_e, aes(x=mT, y=SC), linetype="dotdash")+
   geom_ribbon(data =  dt_plot, aes(x=mT, y=SC, ymin = SC.LCL, ymax = SC.UCL, fill=Period), alpha = 0.1)+
   theme_classic()+
   scale_y_continuous(
      breaks = c(0, 0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1),
      limits = c(0, 1)) +
   geom_hline(yintercept=0.8, linetype=2, color = "black")+
   scale_color_manual(values=c("#F0E442","#E69F00","#0072B2"))+
   scale_fill_manual(values=c("#F0E442","#E69F00","#0072B2"))+
    theme(legend.position = "none",
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          strip.text = element_text(size = 14),
          axis.text.x = element_text(size = 12),
          axis.text.y = element_text(size = 12))+
    facet_wrap(~ Period, nrow = 2)
  return(p)
}

iNEXT_size_cov_53 <- size_vs_coverage(iNEXT53plot)
iNEXT_size_cov_67 <- size_vs_coverage(iNEXT67plot)

ggsave(iNEXT_size_cov_53 , file=paste0(plot_dir, "/Chapter5/iNEXT_size_cov_53.jpg"), width=8, height=8)
ggsave(iNEXT_size_cov_67, file=paste0(plot_dir, "/Chapter5/iNEXT_size_cov_67.jpg"), width=8, height=8)



diversity_vs_size <- function(dt_plot) {
  dt_plot_ref <- subset(dt_plot, dt_plot$Method=="Observed")
  dt_plot_r <- subset(dt_plot, dt_plot$Method=="Rarefaction")
  dt_plot_e <- subset(dt_plot, dt_plot$Method=="Extrapolation")
  
  p <- ggplot(data=dt_plot, aes(x=mT, y=qTD, color=Period)) +
    labs(y="species richness", x="size")+
    geom_point(data=dt_plot_ref, aes(x=mT, y=qTD, color=Period))+
    geom_line(data=dt_plot_r, aes(x=mT, y=qTD, color=Period), linetype="solid", alpha=0.4)+
    geom_line(data=dt_plot_e, aes(x=mT, y=qTD, color=Period), linetype="dotdash")+
    geom_ribbon(data =  dt_plot, aes(x=mT, y=qTD, ymin = qTD.LCL, ymax = qTD.UCL, fill=Period), alpha = 0.1)+
    theme_classic()+
    scale_color_manual(values=c("#F0E442","#E69F00","#0072B2"))+
    scale_fill_manual(values=c("#F0E442","#E69F00","#0072B2"))+
    theme(legend.position = "none",
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          strip.text = element_text(size = 14),
          axis.text.x = element_text(size = 12),
          axis.text.y = element_text(size = 12))+
    facet_wrap(~Period, nrow=2)
    
  return(p)
  
}

iNEXT_size_div_53 <- diversity_vs_size(iNEXT53plot)
iNEXT_size_div_67 <- diversity_vs_size(iNEXT67plot)

ggsave(iNEXT_size_div_53 , file=paste0(plot_dir, "/Chapter5/iNEXT_size_div_53.jpg"), width=8, height=8)
ggsave(iNEXT_size_div_67, file=paste0(plot_dir, "/Chapter5/iNEXT_size_div_67.jpg"), width=8, height=8)

```

<p>&nbsp;</p>
#### **Estimated gamma diversity (Survey data 67):**

```{r, echo=TRUE, results=TRUE}
# SpecPool : ===================================================================
# This chunck uses both the 67 site subset

listSpecpool <- lapply(list_mat67, function(x){specpool(x)}) 
names(listSpecpool) <- c("a) Historical Conservative", "b) Historical Broad", "c) Contemporary Natives + Exotics")

listSpecpool <- do.call(rbind,listSpecpool)
listSpecpool %>%
  kbl() %>%
  kable_styling()

write.csv(round(listSpecpool,2), file=paste0(plot_dir, "/Chapter5/listSpecpool.csv"), row.names = F)

```
<p>&nbsp;</p>

__End of script__
